Nucleotide Diversity Loss on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-11-13 manuscript/manuscript.tex
full edits up to methods: simulations

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can affect levels of genetic diversity. On Y chromosomes with suppressed recombination, selection is expected to eliminate neutral variation and reduce the $N_{e}$ of Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male reproductive success can also reduce Y-linked $N_{e}$, making it difficult to infer the causes of low Y-diversity. Here, we investigate the factors affecting levels of polymorphism within the genome during sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae), a dioecious plant with young sex chromosomes. Strikingly, we find that neutral diversity for genes on the Y is on average \textasciitilde 2.1\% of the value for their homologues on the X, corresponding to a chromosome-wide reduction of \textasciitilde 93\% compared to the neutral expectation. We demonstrate that the magnitude of this diversity loss is inconsistent with a reduced male $N_{e}$ caused by neutral processes including female-biased sex ratios and high variance in male reproductive success. Instead, using forward simulations, we show that the loss of diversity on the Y can be explained by interference among a large number ($\geq$ 800 Kb) of weakly selected mutations. Our results are in agreement with theory on "interference selection" and provide evidence that the aggregate effects of purifying selection over a large number of genetically-linked sites can substantially reduce neutral diversity. Given the recent origin of \textit{R. hastatulus} sex chromosomes (\textasciitilde 15MYA), our results imply that Y chromosome degeneration in the early stages may be largely driven by selective interference rather than by selectively-favoured gene silencing followed by neutral genetic drift.
\end{abstract}

@@

\lettrine[lines=2]{\color{color2}M}{}orphologically distinct sex chromosomes have evolved multiple times independently in both plants and animals \citep{westergaard1958,ohno1967,bull1983,charlesworth1991}. Despite clear biological differences between these kingdoms, X and Y chromosomes in both lineages have undergone similar genetic changes. For example, in both groups the loss of recombination between X and Y chromosomes has been associated with an accumulation of deleterious mutations and a gradual loss of genes from the Y chromosome \citep{hough2014,bergero2015,bachtrog2013NRG}, and in some species, the degeneration of the Y has led to the evolution dosage compensation of the X chromosome \citep{charlesworth1996CB,muyle2012,mank2013sex,papadopulos2015}. The independent evolution of these phenomena in such taxonomically distant species suggest that general evolutionary mechanisms may be involved, but inferring the causes of molecular evolution and patterns of polymorphism in genomic regions that lack recombination is a longstanding challenge for both theoreticians and experimentalists \citep{charlesworth1978,feldman1980evolution,barton1995general,charlesworth1996CB,otto1997deleterious,charlesworth2000degeneration,mcvean2000effects}.
% and reduced genetic variation on Y chromosomes \citep{qiu2010nucleotide, ellegren2011,bachtrog2013NRG}.

One fundamental difference between the X and Y chromosomes is that there are 1/3 as many Y-linked gene copies as X-linked ones in a diploid population. Thus, genes on the Y chromosome are expected to experience an effective population size ($N_{e}$) that is 1/4 that of autosomal genes, whereas genes on the X chromosome should have an $N_{e}$ that is 3/4 that of autosomal genes (assuming an equal number of reproducing females and males). The lowered $N_{e}$ of the Y chromosome implies that that the equilibrium level of neutral polymorphism, which is proportional to the product of $N_{e}$ and the neutral mutation rate, $\mu$, should be lower for Y-linked genes than for their X-linked counterparts. In the absence of recombination, genes on the Y chromosome are also expected to be in strong linkage disequilibrium, making them vulnerable to diversity loss due to selection against strongly deleterious mutations (background selection) and selective sweeps of strongly beneficial mutations (genetic hitchhiking). Furthermore, the build-up of linkage disequilibrium between selected mutations on the Y means that selection will act non-independently across the chromosome, and previous work has shown that such "selective interference" can significantly reduce both the efficacy of selection and the level of neutral variability \citep{fisher1930genetical, muller1964relation, hill1966HReffect, mcvean2000}. These arguments all suggest that non-recombining Y chromosomes should harbor a lower amount of neutral genetic variability than predicted based on the number of Y chromosomes in a population, but the extent to which background selection, genetic hitchhiking, or selective interference should affect chromosome-wide levels of diversity, and the relative importance of these processes in natural populations, is not well-understood.

In addition to reduced $N_{e}$ arising from selection, species with female-biased sex ratios or extensive male-male competition, a high variance in male reproductive success is also predicted reduce the $N_{e}$ experience by genes on the Y chromosome \citep{caballero1995,charlesworth2001,laporte2002,pool2007,ellegren2009}, suggesting that inferences about the effects of selection and interference need to be distinguished from these effects. Because variance in male reproductive success reduces both Y-linked $N_{e}$ and autosomal $N_{e}$ \citep{kimura1964number,nomura2002effective}, evidence for this effect can therefore be obtained by comparing levels of neutral diversity on X and Y chromosomes relative to values on autosomes. For example, high variance in male reproductive success causes a reduction in the Y/A diversity ratio, but an increase in the X/A ratio. Based on such comparisons, studies in humans have suggested that the inflated X/A ratio is attributable to a historical excess of breeding females over males \citep{hammer2008sex} (and see \citep{bustamante2009,hammer2010,cotter2016genetic})

Despite widespread interest in determining the evolutionary factors affecting neutral diversity on sex chromosomes \citep{ellegren2011,bachtrog2013NRG}, we know very little about the influence of either reproductive variance or selection in determining level of diversity on more recently evolved sex chromosomes. The time scales over which these effects are likely to be important is therefore not well understood. In humans, estimates of Y-linked diversity are considerably lower than predicted under neutral models, and simulations have shown that current levels of Y-diversity are consistent with strong purifying selection \citep{Wilsonsayres2014}. However, given that human sex chromosomes evolved from autosomes \textasciitilde 200 million years ago (MYA), is not immediately obvious whether such strong effects of purifying selection on diversity should also apply to sex chromosomes that evolved \textit{do novo} from autosomes over much more recent evolutionary time (e.g., within the last 20 MYA in the case of plants \citep{charlesworth2015plant}). Y-linked diversity loss due to purifying selection might be expected to be lower in such systems due to recombination having been suppressed for a shorter period of time. However, simulations have shown that background selection and genetic hitchhiking have the strongest effects during the earliest stages of sex chromosome evolution, before the Y has lost the majority of its genes \citep{bachtrog2008temporal}. Indeed, this is consistent with studies of Y-degeneration in young plant sex chromosome systems, which have found evidence for rapid genetic degeneration \citep{hough2014,papadopulos2015,charlesworth2016plant}. Despite this, the large number of sites between which recombination is suppressed on a recently evolved Y chromosome means that classic predictions from background selection theory break down (refs), and it is ,...

%In these cases, if Y-degeneration involved widespread gene silencing, as predicted by \citep{orr1998adaptive} and discovered on the neo-Y in \textit{Drosophila mrianada} \citep{zhou2012chromosome}, then selection might in fact play a less important role than theory would predict.
%with the early stages of Y-degeneration instead being primarily driven by gene-silencing followed by neutral genetic drift \citep{bachtrog2013NRG}.

%The strength of this effect is dependent on the rate of recombination between a selected locus and a neighbouring region; essentially, the effect becomes more pronounced with less recombination because of linkage disequilibrium between adjacent loci.

%On the other hand, if gene silencing and Y inactivation occur very early during the process of Y chromosome degeneration, it is possible that the strength of linked selection is minimized rapidly even on relatively young Y chromosomes, and that much degeneration may occur neutrally following the silencing of Y-linked genes (cite backtrog NRG review).

%because recombination suppression was relatively recent on plant Y chromosomes, it is an open question whether weak-HRI has had enough time to have a large effect. Although selective interference (indlucing BGS and Selective Sweeps) is expected to have the greatest effects during the early stages of Y-chromosome evolution \citep{bachtrog2008temporal} the R. hastatulus Y chromosome has undergone significant degeneration and gene loss over a short period of time, suggesting that it may behave more similarly to more ancient Y chromosomes.

%Background selection (purifying selection against deleter- ious alleles) [27] and selective sweeps (positive selection for beneficial variants) [28] both have the effect of reducing nucleotide diversity at linked sites.
%as modeled by \citep{kimura1964number} and \citep{nomura2002effective}.

%evolutionary theory predicts that neutral variability on Y chromosomes with suppressed recombination can be significantly reduced by Hill-Robertson Interference (HRI), which causes a reduction in $N_{e}$ of neutral sites that are genetically linked to sites experiencing selection. The major types of HRI include (i) selection on strongly beneficial mutations ("selective sweeps") \citep{smith1974hitch,aquadro1994} (ii) selection against strongly deleterious mutations) \citep{charlesworth1996background,charlesworth1994effect} \citep{charlesworth1996CB,charlesworth2000degeneration}, and (iii) intereference between weakly selected sites (weak selection Hill-Robertson interference). In all of these processes, linkage  disequilibrium between selected and neutral sites results in a level of neutral variation below the level predicted in the absence of selection.

%Consistent with an important role for Hill-Robertson interference in driving down diversity on the Y chromosome, s

%discussion point?:
%Moreover,

To investigate the factors affecting nucleotide diversity in the early stages of sex chromosome evolution, we analyzed neutral polymorphism levels on X, Y, and autosomal chromosomes in the plant \textit{R. hastatulus }(Polygonaceae). This species is a dioecious annual with highly heteromorphic X and Y chromosomes that are estimated to have evolved  approximately 15 MYA \citep{quesada2011,grabowska2015,navajas2005}, making sex chromosomes in this species over 100 million years younger than the well-studied mammalian sex chromosomes \citep{lahn1999,ross2005dna}. \textit{Rumex hastatulus} has also received particular attention because of the unique occurrence of an intraspecific polymorphism in sex chromosome system, in which both XY and $XY_{1}XY_{2}$ males occur in geographically distinct populations ('chromosomal races') \citep{smith1963mechanism}. The $XY_{1}XY_{2}$ sex chromosome system in this species (the North Carolina race) is thought to have originated through an X-autosome fusion, with the XY system (the Texas race) maintaining the ancestral chromosome complement \citep{smith1964evolving}. Despite the recent origin of sex chromosomes in both races, there is evidence that both Y's have undergone significant gene loss and functional deterioration \citep{hough2014}. Here, to simplify our comparison of polymorphism levels on X, Y, and autosomes, we focus only on the ancestral Y chromosome, which occurs in both sex chromosome races.

Of particular relevance to our study, \textit{R. hastatulus} populations have been found to consistently exhibit female-biased reproductive sex ratios, with a mean sex ratio of $N_{f}/(N_{m}+N_{f})=0.6$ \citep{pickup2013influence}. Female-biased sex ratios in dioecious plants with heteromorphic X and Y chromosomes are not uncommon (\X refs) \citep{hough2013evolutionarily}, and their occurrence in \textit{R. hastatulus} provides an excellent opportunity to study both the demographic and selective factors contributing to sex-linked variability. Using estimates of neutral polymorphism on sex chromosomes and autosomes, together with forward population genetic simulations, our study tests whether sex chromosome evolution in the early stages is driven primarily by neutral processes, or whether linked selection has played a significant role.

@@

\subsection*{Estimating nucleotide diversity on sex chromosomes and autosomes}
For each locus in our analysis, we calculated Watterson’s (1975) estimator of the population parameter $\theta=4N_{e}\mu$, where $N_{e}$ is the effective population size, and $\mu$ is the mutation rate \citep{watterson1975}, using a modified version of the Perl program Polymorphurama \citep{bachtrog2006}. To compare sex-linked and autosomal loci, we calculated the average value of $\theta$, weighted by the number of synonymous sites in each gene (\X Figure 2; Table 1). We obtained 95 percent confidence intervals for our estimates of the X/A and Y/A diversity ratios by bootstrapping per gene using the BCa method \citep{efron1994} implemented in the Boot package in R \citep{canty2012boot}, and calculating X/A and Y/A diversity on each iteration for 20000 replicates each. Bootstrapping was conducted on the final filtered set of 173 sex-linked, and 12355 autosomal genes. Note that the lack of recombination on the Y chromosome implies that assumptions about independence across loci are violated, suggesting that the true uncertainty in our Y/A diversity estimate may be wider than implied by bootstrapping. To address this, we also used a maximum likelihood approach, implemented in a modified version of the MLHKA software \citep{wright2004hka}, to independently estimate a credibility interval for the Y/A ratio (Figure S1). Because of the thousands of genes involved, a likelihood method incorporating divergence to control for heterogeneity in mutation rate was not feasible, as this would require maximizing the likelihood estimate of the mutation rate for each locus independently. Therefore, we assumed no heterogeneity in mutation rate, no recombination between Y-linked genes, and free recombination between autosomal loci. Our model thus had two parameters: $\theta_{autosomal}$ and Y/A. We varied both parameters and evaluated the likelihood for Y/A from 0.001 to 1, and $\theta_{autosomal}$ from 0.001 to 0.01.

%from the Texas race, and separately for the 176 sex-linked and 11349 autosomal genes from the North Carolina race.
%We also tested whether estimates of diversity for the X chromosome calculated from phased sequences from females were consistent with estimates from phased sequences from males. As no significant difference was observed (\X Figure? Table?), we report only results from females.

%Finally, we tested for population substructure within and between the two sex chromosome races to control for the possibility that hidden substructure in our sampled populations could affect estimates of diversity. To do this, we constructed neighbor-joining trees of sex-linked sequences from all populations in our study. The analysis was conducted on an alignment of the X- and Y-linked genes from \textit{R. hastatulus}, with orthologous autosomal sequences from the non-dioecious but closely related outgroup species \textit{R. bucephalophorus} used to root the tree (Figure S2). We used the Neighbor-Joining method \citep{saitou1987neighbor}, with evolutionary distances computed using Maximum Composite Likelihood \citep{tamura2011mega5}. The inferred trees revealed strong support for Y-linked genes from the $XY_{1}XY_{2}$ ("North Carolina") race being paraphyletic (98\% bootstrap support), with samples from two populations (hereafter the "SC sub-clade") forming a monophyletic group, and samples from Florida and Georgia (hereafter the "FL sub-clade") also forming a monophyletic group, that was more closely related to the XY ("Texas") race (Figure S2). In light of this evidence for significant population substructure, and to consider between-population differences, we estimated X and Y diversity in each of the three sub-clades (SC, FL, TX) separately. For model-based analyses, we focused on the population sample from the Texas XY race, given that this reflects the presumed ancestral karyotype and has the largest sample size.

\subsection*{Neutral predictions and the effect of sex ratio bias on diversity}
To test whether the levels of diversity we observed on X, Y and autosomal chromosomes could be explained by the occurrence of biased reproductive sex ratios or high variance in reproductive success, we compared our empirical diversity estimates to predictions from a neutral model that considers these two effects \citep{kimura1964number}. In particular, we calculated the effect of variance in reproductive success on $N_{e}$ as:

@@

where $k_{m}$ and $V_{k_{m}}$ are the mean and variance in male offspring number, respectively, and $\overline{k}$ is the total mean number of progeny \citep{kimura1964number}. Given the expected $N_{m}$ and $N_{f}$ for varying levels $V_{k_{m}}$, the expected $N_{e}$ for autosomes and sex chromosomes was then calculated using equations from \citep{wright1931evolution}:

@@

Note that with equal sex ratios, $N_{e_{X}}/N_{e_{A}} = 0.75$, and $N_{e_{Y}}/N_{e_{A}} = 0.25$. Figure 1 shows these expected X/A and Y/A ratios as a function of the sex ratio. We tested the fit of our empirical diversity ratios to these predictions (shown in Figure 2) using the sex ratio of 0.6 as estimated in \textit{R. hastatulus}.

%To consider the combined effects on diversity of our empirically estimated female-biased sex ratio ($(N_{f}$/$N_{m}+N_{f})$=$0.6$ together with high variance in male reproductive success, we calculated the nuetrally-expected X/A and Y/A ratios under a 0.6 sex ratio across a range of values for $C^2_{d{m}}$: 0 (no variance), 1 (Poisson distribution of offspring), 2 (Poisson*2), and 3 (Poisson*3).
%\begin{equation}
%N_{e{A}} = \frac{4N_{m}N_{f}}{N_{m}(1+C^2_{f})+N_{f}(1+C^2_{m})}\label{eq:NeA}
%\end{equation}

%where $N_{m}$ and $N_{f}$ are effective population sizes for males and females, respectively, and $C^2_{i}$ is the coefficient of variation in mating success for the $i$th sex, defined as the ratio of the standard deviation to the mean number of mates $\sigma_{i}/\mu_{d{i}}$. Similarly, for X-linked loci, where females contain two-thirds and males one-third of the alleles:

%\begin{equation}
%N_{e{X}} = \frac{9N_{m}N_{f}}{4N_{m}(1+C^2_{f})+2N_{f}(1+C^2_{m})}\label{eq:NeX}
%\end{equation}

%With no variation in mating success in either sex ($C^2_{d{m}}=C^2_{d{f}}=0$), equations 1 and 2 reduced to the familiar expressions from \citep{wright1931evolution}:

%\begin{equation}
%\label{eq:NeY}
%\end{equation}

%Because the $N_{e}$ for the Y chromosome depends on the male effective population size, $N_{m}$, the effect of variance in reproductive success on $N_{Y}$ can be modeled using the approach of Crow and Kimura (refs; and see ) and $N_{m}$ is given by:

%\begin{equation}
%N_{e{X}} = \frac{N(k-1)}{k-1}
%\end{equation}

%\begin{equation}
%\label{eq:NeY}
%\end{equation}

%Using these predictions, we calculated the expected X/A and Y/A ratios of diversity for the estimated \textit{R. hastatulus} sex ratio (N_{f}/$N_{m}+N_{f})=0.6$ \citep{pickup2013influence}, assuming that the level of neutral polymorphism is given by the product of the mutation rate and the effective population size, $\theta=4N_{e}\mu$ \citep{watterson1975}. Sex-specific variance in reproductive success can lead to departures from this prediction, and so we also evaluated whether any ratio of male- to female- effective population sizes could explain our observed combination of X/A and Y/A diversity.

\begin{figure}[htbp]
\centering
\noindent
\includegraphics[width=\linewidth]{figure1.jpg}
\caption{The relation between relative effective population size at equilibrium and sex ratio bias for genes on autosomes (\textbf{A}), the X chromosome (\textbf{B}), and the normalized X/A and Y/A ratios (\textbf{C}). The sex ratio is shown as the proportion of females, $N_{f}/(N_{f}+N_{m})$, plotted against $N_{e}/N$, where $N=N_{m}+N_{f}$ and the $N_{e}$'s' for sex chromosomes and autosomes are given in equations 2-4. Dotted curves show predictions when both males and females produce Poisson-distributed offspring numbers ($\overline{k}=2$). Solid curves correspond to increasing levels of variance in male reproductive success, ranging from Poisson/2 (top solid curve in panels A and B) to 3*Poisson (bottom most curve). Assuming $\theta=4N_{e}\mu$ and an equal neutral mutation rates among genes, these predicted $N_{e}$'s were used to generate our null neutral predictions for the X/A and Y/A ratios of diversity.
}
\label{fig:spectrum}
\end{figure}

>>>>

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can affect levels of genetic diversity. On Y chromosomes with suppressed recombination, selection is expected to eliminate neutral variation and reduce the $N_{e}$ of Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male reproductive success can also reduce Y-linked $N_{e}$, making it difficult to infer the causes of low Y-diversity. Here, we investigate the factors affecting levels of polymorphism within the genome during sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae), a dioecious plant with young sex chromosomes. Strikingly, we find that neutral diversity for genes on the Y is on average \textasciitilde 2.1\% of the value for their homologues on the X, corresponding to a chromosome-wide reduction of \textasciitilde 93\% compared to the neutral expectation. We demonstrate that the magnitude of this diversity loss is inconsistent with a reduced male $N_{e}$ caused by neutral processes including female-biased sex ratios and high variance in male reproductive success. Instead, using forward simulations, we show that the loss of diversity on the Y can be explained by interference among a large number ($\geq$ 800 Kb) of weakly selected mutations. Our results are in agreement with theory on "interference selection", and provide evidence that the effects of purifying selection over a large number of genetically-linked sites can substantially reduce neutral diversity. Given the recent origin of \textit{R. hastatulus} sex chromosomes (\textasciitilde 15MYA), our results imply that Y chromosome degeneration in the early stages may be largely driven by interference rather than positive selection for gene silencing followed by neutral genetic drift.
\end{abstract}

@@

\lettrine[lines=2]{\color{color2}M}{}orphologically distinct sex chromosomes have evolved multiple times independently in both plants and animals \citep{westergaard1958,ohno1967,bull1983,charlesworth1991,charlesworth2015plant}. Despite clear biological differences between these kingdoms, X and Y chromosomes in both lineages have undergone similar genetic changes. For example, in both groups the loss of recombination between X and Y chromosomes has been associated with an accumulation of deleterious mutations and a gradual loss of genes from the Y chromosome \citep{hough2014,bergero2015,bachtrog2013NRG}, and in some species, the degeneration of the Y has led to the evolution dosage compensation of the X chromosome \citep{charlesworth1996CB,muyle2012,mank2013sex,papadopulos2015}. The independent evolution of these phenomena in such taxonomically distant species suggest that general evolutionary mechanisms may be involved, but inferring the causes of molecular evolution and patterns of polymorphism in genomic regions that lack recombination is a longstanding challenge for both theoreticians and experimentalists \citep{charlesworth1978,feldman1980evolution,barton1995general,charlesworth1996CB,otto1997deleterious,charlesworth2000degeneration,mcvean2000effects}.

One fundamental difference between the X and Y chromosomes is that there are 1/3 as many Y-linked gene copies as X-linked ones in a diploid population. Thus, genes on the Y chromosome are expected to experience an effective population size ($N_{e}$) that is 1/4 that of autosomal genes, whereas the  $N_{e}$ for genes on the X chromosome should be 3/4 that of autosomal genes (assuming an equal number of reproducing females and males). The lowered $N_{e}$ of the Y chromosome implies that that the equilibrium level of neutral polymorphism - proportional to the product of $N_{e}$ and the neutral mutation rate, $\mu$ - should be lower for Y-linked genes than for their X-linked counterparts. In the absence of recombination, genes on the Y chromosome are also expected to be in strong linkage disequilibrium, making them vulnerable to diversity loss due to selection against strongly deleterious mutations (background selection) and selective sweeps of strongly beneficial mutations (genetic hitchhiking). Furthermore, the build-up of linkage disequilibrium between selected mutations on the Y means that selection will act non-independently across the chromosome such that selection at a focal site may "interfere" with selection at sites with which it is linked. Originally developed by Hill & Robertson (1966), a large body of work has now shown that such "selective interference" can substantially reduce both the efficacy of selection and the level of neutral variability \citep{fisher1930genetical, muller1964relation, hill1966HReffect, mcvean2000}. These arguments all suggest that non-recombining Y chromosomes should harbor a lower amount of neutral genetic variability than predicted based on the number of Y chromosomes in a population, but the extent to which background selection, genetic hitchhiking, or selective interference have affected chromosome-wide levels of diversity in natural populations, and the relative importance of these processes, is not well-understood.

In addition to reduced diversity arising from selection, in species with female-biased sex ratios or extensive male-male competition, high variance in male reproductive success is also expected reduce the $N_{e}$ experience by genes on the Y chromosome \citep{caballero1995,charlesworth2001,laporte2002,pool2007,ellegren2009}, suggesting that inferences about the effects positive of purifying selection need to be distinguished from these neutral processes. Because variance in male reproductive success reduces both Y-linked $N_{e}$ and autosomal $N_{e}$ \citep{kimura1964number,nomura2002effective}, evidence for this can therefore be obtained by comparing levels of neutral diversity on X and Y chromosomes relative to values on autosomes. For example, high variance in male reproductive success causes a reduction in the Y/A diversity ratio, but an increase in the X/A ratio. Based on such comparisons, studies in humans, for example, have suggested that the inflated X/A ratio is attributable to a historical excess of breeding females over males \citep{hammer2008sex} (and see \citep{bustamante2009,hammer2010,cotter2016genetic}).

Despite widespread interest in determining the evolutionary factors affecting neutral diversity on sex chromosomes \citep{ellegren2011,bachtrog2013NRG}, we know very little about the influence of either sex ratio variation or linked selection in determining levels of diversity on more recently evolved sex chromosomes. The time scales over which these different effects are likely to be important is therefore not well understood. In humans, estimates of Y-linked diversity are considerably lower than predicted under neutral models, and simulations suggest that levels of diversity are consistent with strong purifying selection \citep{Wilsonsayres2014}. However, given that human sex chromosomes evolved from autosomes \textasciitilde 200 million years ago (MYA), is not immediately obvious whether purifying selection might have such strong effects on Y chromosomes that evolved \textit{do novo} from autosomes over much more recent evolutionary time (e.g., within the last 20 MYA in the case of plants \citep{charlesworth2015plant}). Y-linked diversity loss might on the one hand be expected to be lower in younger systems due to a shorter history of recombination suppression. On the other hand, simulations of strong selection models (background selection and genetic hitchhiking) suggest that these processes may have the greatest effects during the earliest stages of sex chromosome evolution, before the Y has lost many of its genes \citep{bachtrog2008temporal}. Moreover, work by \citep{KaiserCharlesworth} and \citep{good2014genetic} have provided evidence that even weak purifying selection, if acting non-independently across large number of sites, can generate substantial deviations from neutrality, whereas classic background selection theory breaks down in such cases. Given that a large number of sites are likely to be in linkage disequilibrium on a recently evolved Y chromosome, such "interference selection" \textit{sensu} \citep{good2014genetic} is a good candidate model for exploring the evolutionary dynamics of young plant sex chromosomes.

To investigate the factors affecting nucleotide diversity in the early stages of sex chromosome evolution, we analyzed neutral polymorphism levels on X, Y, and autosomal chromosomes in the plant \textit{R. hastatulus }(Polygonaceae). This species is a dioecious annual with heteromorphic X and Y chromosomes that evolved approximately 15 MYA \citep{quesada2011,grabowska2015,navajas2005}, making Y chromosomes in this species over 100 million years younger than the highly degenerated Y chromosomes in mammals \citep{lahn1999,ross2005dna}. \textit{Rumex hastatulus} has also received particular attention because of the occurrence of an intraspecific polymorphism in sex chromosome system, in which both XY and $XY_{1}XY_{2}$ males occur in geographically distinct populations ("chromosomal races") \citep{smith1963mechanism}. The $XY_{1}XY_{2}$ sex chromosome system in this species is thought to have originated through an X-autosome fusion, with the XY system maintaining the ancestral chromosome complement \citep{smith1964evolving}. Despite the recent origin of sex chromosomes in both races, there is evidence that both Y's have undergone significant gene loss and functional deterioration \citep{hough2014}. Here, to simplify our comparison of polymorphism levels on X, Y, and autosomes, we focus only on the ancestral Y chromosome, which occurs in both sex chromosome races.

Of particular importance for our study, \textit{R. hastatulus} populations have been found to consistently exhibit female-biased reproductive sex ratios, with a mean sex ratio of $N_{f}/(N_{m}+N_{f})=0.6$ \citep{pickup2013influence}. Female-biased sex ratios are not uncommon in dioecious plants with heteromorphic X and Y chromosomes \citep{field2013comparative,hough2013evolutionarily}, and their occurrence in \textit{R. hastatulus} provides an excellent opportunity to study both the demographic and selective factors contributing to sex-linked variability.

@@

\subsection*{Estimating nucleotide diversity on sex chromosomes and autosomes}
For each locus in our analysis, we calculated Watterson’s (1975) estimator of the population parameter $\theta=4N_{e}\mu$, where $N_{e}$ is the effective population size, and $\mu$ is the mutation rate \citep{watterson1975}, using a modified version of the Perl program Polymorphurama \citep{bachtrog2006}. To compare sex-linked and autosomal loci, we calculated the average value of $\theta$, weighted by the number of synonymous sites in each gene (\X Figure 2). We obtained 95\% confidence intervals for X/A and Y/A ratios by bootstrapping per gene using the BCa method \citep{efron1994} implemented in the Boot package in R \citep{canty2012boot}, and calculating X/A and Y/A on each iteration for 20000 replicates each. Bootstrapping was conducted on the final filtered set of 173 sex-linked, and 12355 autosomal genes. Note that the lack of recombination on the Y chromosome implies that statistical assumptions about independence across loci are violated, suggesting that the true uncertainty in the estimate Y/A may be wider than implied by bootstrapping results. To address this, we also used a maximum likelihood approach, implemented in a modified version of the MLHKA software \citep{wright2004hka}, to independently estimate a credibility interval for the Y/A ratio (Figure S1). Because of the thousands of genes involved, a likelihood method incorporating divergence to control for heterogeneity in mutation rate was not feasible, as this would require maximizing the likelihood estimate of the mutation rate for each locus independently. Therefore, we assumed no heterogeneity in mutation rate, no recombination between Y-linked genes, and free recombination between autosomal loci. Our model thus had two parameters: $\theta_{autosomal}$ and Y/A. We varied both parameters and evaluated the likelihood for Y/A from 0.001 to 1, and $\theta_{autosomal}$ from 0.001 to 0.01.

\subsection*{Neutral predictions and the effect of sex ratio bias on diversity}
To test whether the levels of diversity on X, Y and autosomal chromosomes could be explained by neutral processes, we compared our diversity estimates to predictions from a model that incorporates variance of progeny number for males and females, and gives the predicted $N_{e}$ for males and females when this variance is either smaller or larger than Poisson \citep{kimura1964number, hedrick2011genetics}. As we were primarily interested in the possible influence of variance in male reproductive success on Y-linked $N_{e}$, and the parameter space within which such a model might generate Y/A and X/A ratios consistent with our data, we ignored non-poission variance in female offspring, and calculated the male $N_{e}$ as a function of variance in offspring number as:

@@

where $k_{m}$ and $V_{k_{m}}$ are the mean and variance in male offspring number, respectively, in a population of $N_{m}$ reproducing males and $\overline{k}$ is the total mean number of progeny of males and females \citep{kimura1964number}. Given the expected values for $N_e_{m}$ across a range of values for $V_{k_{m}}$ and $N_{m}$, the corresponding expectation for the $N_{e}$ of autosomes and sex chromosomes at equilibrium were calculated as:

@@

following \citep{wright1931evolution}. Note that with equal sex ratios, and Poisson distributed offspring numbers ($V_{k_{m}} = \overline{k}$), then $N_{e_{X}}/N_{e_{A}} = 0.75$, and $N_{e_{Y}}/N_{e_{A}} = 0.25$. Figure 1 shows these predictions as a function of the sex ratio for a range values of $V_{k_{m}}$, and we tested the fit of our empirical diversity ratios to these predictions (shown in Figure 2) assuming the estimated \textit{R. hastatulus} sex ratio of 0.6 \citep{pickup2013influence}.

\begin{figure}[htbp]
\centering
\noindent
\includegraphics[width=\linewidth]{figure1.jpg}
\caption{The relation between effective population size and sex ratio bias for genes on autosomes (\textbf{A}) and the X chromosome (\textbf{B}), and corresponding normalized X/A and Y/A ratios (\textbf{C}). The sex ratio is shown as the proportion of females, $N_{f}/(N_{f}+N_{m})$, plotted against $N_{e}/N$, where $N=N_{m}+N_{f}$. Dotted curves show equilibrium predictions when both males and females produce Poisson-distributed offspring numbers and ($V_{k}$=$\overline{k}=2$). Solid curves correspond to increasing values for $V_{k_{m}}$, the variance in male offspring number, ranging from Poisson/2 (top solid curve in panels A and B) to 3*Poisson (bottom most curve). For (\textbf{C}) we assumed equal neutral mutation rates among genes and calculated the expected neutral diversity as $\theta=4N_{e}\mu$.
}
\label{fig:spectrum}
\end{figure}
